---
title: "Income and Renewable Heating Adoption in Switzerland"
author: "Yannik Zimmermann & Birk Diener"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
library(scales)
library(sf)
library(BFS)
library(maps)

```

## Motivation

Switzerland has committed to achieving net-zero greenhouse gas emissions by 2050, with the heating sector representing one of the largest remaining sources of fossil fuel consumption. Transitioning residential buildings away from oil and gas heating towards renewable alternatives such as heat pumps, solar thermal systems, and biomass is therefore a critical lever of the energy transition. However, adoption rates of renewable heating systems vary significantly across Swiss municipalities (bfs 2024). This project investigates a potential explanation for this heterogenous adoption pattern, namely whether a statistically measurable relationship exists between community-level income and the share of renewable heating systems installed.

## Project Aim

This project aims to quantify the relationship between median income at the municipality level and the share of buildings equipped with renewable heating systems across Swiss municipalities. The analysis will address two core research questions: 

- Is there a statistically significant relationship between municipal-level financial capacity, measured through taxable income, and the share of residential buildings equipped with renewable heating systems across Swiss municipalities?
- To what extent is this relationship robust to structural characteristics of the building stock, namely the share of single-family houses and the age of the building stock, and to cantonal differences in subsidy programmes?

The analysis draws on the Federal Building and Dwellings Register (GWR) and ESTV municipal tax data. Given known data quality limitations in the GWR heating system entries, a systematic data recency assessment is integrated into the analysis prior to modelling. The findings will be visualised through choropleth maps and scatter plots to make spatial and socioeconomic patterns legible for a broader audience.

## Data Sources

Two datasets are used for this analysis, both openly available and joinable at the Swiss municipality level (BFS-Gemeindecode). The first is the Federal Building and Dwellings register (GWR), which records the primary heating energy, building category, and year of construction for every residential building in Switzerland. Key variables used in this analysis are the heating system type (GWAERZH1), the energy source (GENH1), the year of construction (GBAUJ), the information source of the heating entry (GWAERSCEH1), the last update date of the heating entry (GWAERDATH1), and the municipality code (GGDENR).

While the general availability of data to answer the research question is good, there is a significant challenge regarding the reliability of the heating system entries. The last comprehensive national update of this information took place as part of the Volkszählung 2000. Since then, municipal building departments are responsible for updating entries, primarily through building permit applications. As a result, data quality varies across municipalities, and heating systems installed without a building permit — such as most external air-source heat pumps — may not be recorded. We investigate this aspect further in the data preparation chapter.  

The second dataset is the income and wealth statistics from the Federal Tax Administration (ESTV), which provides median taxable income per municipality at the level of natural persons. It is published annually with a lag of approximately two years.

## Approach

The analysis proceeds in four steps: First, GWR building data is aggregated to the municipality level to compute the share of residential buildings equipped with a renewable heating system as a fraction of all buildings with a recorded heating system entry. Renewable heating is defined as heat pump, solar thermal, or biomass combined with energy source. District heating is excluded from the primary definition as its adoption is structurally determined by local infrastructure rather than individual investment decisions, but is included in a sensitivity analysis. The share of pre-1980 buildings per municipality serves as the building age control variable. 

Second, the municipal renewable heating share is joined to the ESTV income data using the BFS municipality code. Third, regression models are estimated using a stepwise approach: a baseline model with median income as the sole predictor, followed by successive addition of building structure controls (share of single-family houses, share of pre-1980 buildings), sociodemographic controls (municipal type, log population size), and cantonal fixed effects to absorb unobserved variation in subsidy programmes and building regulations. The municipal share of low-quality heating entries is included as a control variable throughout all model specifications. The stability of the income coefficient across specifications is reported. 

Fourth, findings are visualized through choropleth maps showing the spatial distribution of renewable heating adoption and median income respectively.

## Data Preparation
Before proceeding to the analysis, the GWR building data is prepared and aggregated to the municipal level. The preparation process includes filtering for residential buildings, flagging renewable heating systems, andÍ assessing the quality of the heating system entries.

Residential buildings are identified using two GWR variables. First, GKAT (building category) is restricted to codes 1020 (exclusively residential) and 1030 (residential with secondary use), excluding buildings without mostly residential use. Second, GSTAT (building status) is restricted to code 1004, retaining only buildings recorded as currently existing and excluding buildings under construction, demolished, or otherwise inactive. Together, these two filters define the analytical population of existing residential buildings on which all subsequent measures are computed.

Heating systems are categorised based on the energy source variable (GENH1) across ten mutually exclusive groups: no heating system (7500), air-source heat pump (7501), ground- and water-source heat pump (7510–7513), gas (7520), oil (7530), wood and biomass (7540–7543), waste heat (7550), electricity (7560), solar thermal (7570), district heating (7580–7582), and unspecified or other (7598–7599). A building is flagged as renewable if its primary energy source is a heat pump, solar thermal, wood or biomass, waste heat, or recorded as having no heating system. The latter category is included on the grounds that buildings without an active heating system generate no fossil fuel consumption from space heating, making their exclusion from the renewable count methodologically conservative. District heating is retained as a separate flag for sensitivity analysis, as its adoption depends on local infrastructure availability rather than individual investment decisions.

```{r load gwr data}

## 1. Read the GWR building data --------------------------------------------
# The file is tab-separated even though it ends in .csv. We only read the
# columns we actually use (the full file has ~50 columns and ~3.3 M rows).

buildings <- read_tsv(
  here("gwr", "gebaeude_batiment_edificio.csv"),
  col_select = c(GGDENR, GGDENAME, GDEKT,
                 GKAT, GKLAS, GSTAT, GBAUJ,
                 GWAERZH1, GENH1, GWAERSCEH1, GWAERDATH1),
  na = c("", "NA"),
  show_col_types = FALSE
)

## 2. Keep only existing residential buildings -----------------------------
# GKAT: 1020 = exclusively residential, 1030 = residential with side use
# GSTAT 1004 = the building actually exists today.

residential <- buildings %>%
  filter(GKAT %in% c(1020, 1030), GSTAT == 1004)


## 3. Flag each building --------------------------------------------------
# Renewable heating per the project brief:

residential <- residential %>%
  mutate(
    # ------------------------------------------------------------------
    # 1. Heizsystem-Kategorien — GENH1
    # ------------------------------------------------------------------
    is_no_heatsystem = GENH1 == 7500,
    is_hp_air        = GENH1 == 7501,
    is_hp_ground     = GENH1 %in% c(7510, 7511, 7512, 7513),
    is_gas           = GENH1 == 7520,
    is_oil           = GENH1 == 7530,
    is_wood          = GENH1 %in% c(7540, 7541, 7542, 7543),
    is_waste_heat    = GENH1 == 7550,
    is_electric      = GENH1 == 7560,
    is_solar         = GENH1 == 7570,
    is_district      = GENH1 %in% c(7580, 7581, 7582),
    is_other         = GENH1 %in% c(7598, 7599),

    # Aggregierte Heizsystem-Flags
    is_heatpump      = is_hp_air | is_hp_ground,
    is_renewable     = is_heatpump | is_wood | is_solar |
                       is_waste_heat | is_no_heatsystem,

    # ------------------------------------------------------------------
    # 2. Gebäudestruktur
    # ------------------------------------------------------------------
    is_efh           = GKLAS == 1110,

    # ------------------------------------------------------------------
    # 3. Gebäudealter — Flags und Faktorvariable
    # ------------------------------------------------------------------
    is_pre1945       = !is.na(GBAUJ) & GBAUJ < 1945,
    is_postwar       = !is.na(GBAUJ) & GBAUJ >= 1945 & GBAUJ < 1981,
    is_transition    = !is.na(GBAUJ) & GBAUJ >= 1981 & GBAUJ < 2001,
    is_modern        = !is.na(GBAUJ) & GBAUJ >= 2001 & GBAUJ < 2021,
    is_new           = !is.na(GBAUJ) & GBAUJ >= 2021,
    is_age_unknown   = is.na(GBAUJ),

    building_epoch   = factor(
      case_when(
        is.na(GBAUJ)   ~ "unknown",
        GBAUJ < 1945   ~ "pre1945",
        GBAUJ < 1981   ~ "postwar",
        GBAUJ < 2001   ~ "transition",
        GBAUJ < 2021   ~ "modern",
        TRUE           ~ "new"
      ),
      levels = c("pre1945", "postwar", "transition", "modern", "new", "unknown")
    ),

    # ------------------------------------------------------------------
    # 4. Datenqualität
    # ------------------------------------------------------------------
    is_lowqual       = GWAERSCEH1 == 860 | is.na(GWAERDATH1)
  )

residential <- residential %>%
  mutate(across(starts_with("is_"), ~ coalesce(.x, FALSE)))

## 5. Aggregate to the municipality ---------------------------------------
# Renewable share is defined over buildings with a heating entry, so the
# denominator is sum(!is.na(GWAERZH1)), not n().

muni <- residential %>%
  group_by(bfs_id = GGDENR, municipality = GGDENAME, canton = GDEKT) %>%
  summarise(
    # ------------------------------------------------------------------
    # 1. Grundzahlen
    # ------------------------------------------------------------------
    n_buildings          = n(),
    n_missing_heating    = sum(is.na(GENH1)),

    # ------------------------------------------------------------------
    # 2. Heizsystem-Anteile (Nenner: alle Gebäude)
    # ------------------------------------------------------------------
    share_renewable      = sum(is_renewable)  / n(),
    share_heatpump       = sum(is_heatpump)   / n(),
    share_hp_air         = sum(is_hp_air)     / n(),
    share_hp_ground      = sum(is_hp_ground)  / n(),
    share_solar          = sum(is_solar)      / n(),
    share_wood           = sum(is_wood)       / n(),
    share_waste_heat     = sum(is_waste_heat) / n(),
    share_oil            = sum(is_oil)        / n(),
    share_gas            = sum(is_gas)        / n(),
    share_electric       = sum(is_electric)   / n(),
    share_district       = sum(is_district)   / n(),
    share_other          = sum(is_other)      / n(),

    # ------------------------------------------------------------------
    # 3. Gebäudestruktur (Nenner: alle Gebäude)
    # ------------------------------------------------------------------
    share_efh            = sum(is_efh)        / n(),

    # ------------------------------------------------------------------
    # 4. Gebäudealter (Nenner: alle Gebäude)
    # ------------------------------------------------------------------
    share_pre1945        = sum(is_pre1945)    / n(),
    share_postwar        = sum(is_postwar)    / n(),
    share_transition     = sum(is_transition) / n(),
    share_modern         = sum(is_modern)     / n(),
    share_new            = sum(is_new)        / n(),
    share_age_unknown    = sum(is_age_unknown) / n(),

    # ------------------------------------------------------------------
    # 5. Datenqualität (Nenner: alle Gebäude)
    # ------------------------------------------------------------------
    share_lowqual        = sum(is_lowqual)    / n(),

    .groups = "drop"
  )
```


Unfortunately, the heating system entries in the GWR are of limited reliability. The last comprehensive national update of this information took place as part of the Volkszählung 2000, and since then, municipal building departments are responsible for updating entries, primarily through building permit applications. As a result, data quality varies across municipalities, and heating systems installed without a building permit — such as most external air-source heat pumps — may not be recorded.
This is visible in Figure 1, which plots the most recent update year of the heating entry per building: approximately 600,000 residential buildings have not had their heating entry updated since 2000, and a further 1.08 million buildings have no update date recorded at all. Together these two groups represent approximately 50% of the 3.3 million buildings in the GWR.

```{r gwr data quality}

## 4. Quick data-quality check --------------------------------------------
# When was the heating entry last updated? The brief notes this is a problem,
# many entries still go back to the 2000 census.

residential %>%
  mutate(year = as.integer(substr(GWAERDATH1, 1, 4))) %>%
  filter(!is.na(year), year >= 1990) %>%
  ggplot(aes(year)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Latest update of the GWR heating entry",
       x = NULL, y = "Number of buildings") +
  theme_minimal()

```

Figure 1: Distribution of GWR heating entry update years for residential buildings in Switzerland

*Note: The figure shows the number of residential buildings by the year of their most recent heating system entry update (GWAERDATH1). Buildings with no recorded update date (n = 1,083,361) are excluded from the plot. The spike in 2001 reflects entries sourced from the Volkszählung 2000 (GWAERSCEH1 = 860), which represent the last comprehensive national update of heating system data. The subsequent increase from 2019 onwards reflects growing administrative activity following the 2017 revision of the GWR ordinance, which made heating data recording mandatory for new building permits.*

To account for this, each building is assigned a binary low-quality flag if its heating entry either has no recorded update date or is explicitly sourced from the Volkszählung 2000 (GWAERSCEH1 = 860). Aggregated to the municipal level, this produces a share of low-quality heating entries per municipality, which is used as a control variable in the regression models. The median municipality has 58% of its heating entries flagged as low quality, with an interquartile range of 48.5% to 69.3%, indicating that the data quality challenge is structural rather than isolated. Results should therefore be interpreted as conservative estimates: low-quality entries are likely concentrated in older building stock that is disproportionately fossil-heated, meaning the measured renewable heating shares likely underestimate true adoption rates, and any detected income effect is likely attenuated by this noise rather than inflated.


```{r load estv data and join to gwr}

estv_file <- here("estv", "statistik-dbst-np-kennzahlen-mit-2022.xlsx")

income <- read_excel(
  estv_file,
  sheet = "Gemeinden - Communes",
  na    = c("", "NA", "-")
) %>%
  select(bfs_id = 3, median_income = median_steink) %>%
  filter(!is.na(bfs_id))

muni_joined <- muni %>%
  left_join(income, by = "bfs_id")

muni_analysis <- muni_joined %>%
  filter(!is.na(median_income))

```

Of the 2,110 municipalities in the analytical sample, 12 could not be matched to ESTV income data, most likely due to municipal mergers where GWR and ESTV use different reference dates for municipality codes. These municipalities are excluded from the regression analysis, leaving a final sample of 2,098 municipalities.

```{r load additional municipal data}

## STATPOP — Einwohnerzahl pro Gemeinde
statpop <- read_delim(
  here("gwr", "STATPOP2024_GMDE.csv"),
  delim = ";",
  show_col_types = FALSE
) %>%
  select(bfs_id = GMDE, pop_total = BBTOT)

## Raumgliederungen — Gemeindetypologie
raumgliederung <- read_excel(
  here("gwr", "Raumgliederungen.xlsx"),
  skip = 2,
  col_names = c("bfs_id", "muni_name", "canton_id", "canton",
                "district_id", "district", "typology_3",
                "typology_9", "typology_25")
) %>%
  filter(!is.na(bfs_id)) %>%
  select(bfs_id, typology_3, typology_9, typology_25) %>%
  mutate(
    bfs_id     = as.numeric(bfs_id),
    typology_3 = as.integer(typology_3),
    typology_9 = as.integer(typology_9),
    typology_25 = as.integer(typology_25)
  )

muni_analysis <- muni_analysis %>%
  left_join(statpop,       by = "bfs_id") %>%
  left_join(raumgliederung, by = "bfs_id") %>%
  mutate(log_pop = log(pop_total))


```

## Analysis

```{r eda univariate statistics}

## ── EDA ──────────────────────────────────────────────────────────────────────

## 1. Univariate Statistiken ---------------------------------------------------
muni_analysis %>%
  select(share_renewable, median_income, share_efh,
         share_lowqual, log_pop) %>%
  summary()

```

### Univariate Statistics

The dependent variable share_renewable is right-skewed (Median 25.5%, Mean 28.4%, IQR 16.2–37.5%). The majority of municipalities fall between a 15% and 35% share of renewable heating, while a long right-hand tail extends up to 93.6%. The distribution is not skewed enough to warrant a log transformation but should be treated as a proportion during model interpretation.

Median income (median_income) is approximately symmetrically distributed (Median 59,700 CHF, IQR 54,800–65,688 CHF) with a slight right-hand tail reaching up to 115,300 CHF – driven by high-income, tax-advantaged municipalities, primarily in Zug and Schwyz. A log transformation is plausible for the regression here, as the distribution at the lower end (from approx. 30,000 CHF downwards) exhibits stronger left-skewness. The minimum of 6,900 CHF is likely due to isolated micro-municipalities with unreliable FTA (ESTV) entries.

share_efh (share of single-family homes) is now valid, with a mean of 67.9%, a median of 70.0%, and an IQR of 60.7–76.9%. This aligns well with the FSO (BFS) figure of approximately 64% single-family homes at the building level; the unweighted municipal mean is slightly higher due to the large number of small, rural municipalities. The minimum of 11.5% likely represents a major city like Zurich or Geneva, while the maximum of 97.5% corresponds to a small mountain municipality.

The share of low-quality GWR heating entries (share_lowqual, Median 31.8%, IQR 18.3–51.0%) confirms the challenge outlined in the data description: one-third of the entries are outdated or undated. The wide dispersion indicates that using this share as a control variable is sensible, as it absorbs measurement noise at the municipal level.

```{r eda correlation matrix}

## 2. Korrelationsmatrix -------------------------------------------------------
muni_analysis %>%
  select(share_renewable, median_income, share_efh,
         share_pre1945, share_lowqual, log_pop) %>%
  cor(use = "complete.obs") %>%
  round(2)

```

### Correlation Matrix

The most striking finding in the correlation matrix is the negative correlation between median income and the share of renewable heating (r = –0.36). This contradicts the intuitive expectation that higher purchasing power facilitates the adoption of renewable heating systems. This effect is explained by the urban concentration of high-income municipalities: due to structural factors, urban municipalities have lower renewable shares—gas heating and district heating (excluded here) dominate—but above-average incomes.

Also relevant is the negative correlation between municipality size (log_pop) and the renewable share (r = –0.39). Larger municipalities rely more heavily on fossil fuels. Since income and population size are weakly positively correlated (r = 0.13), potential multicollinearity exists, though it is non-critical at this level.

share_pre1945 shows virtually no correlation with share_renewable (r = –0.04), which is surprising at first glance, as older buildings typically have lower renovation rates. A possible explanation: in municipalities with a high share of old buildings, the GWR data entries are also older, potentially biasing the measured renewable rate upward (outdated entries may fail to classify electric heating as fossil-fuel-based, for example). The low correlation of share_lowqual with all other variables (max. |r| = 0.15) suggests that data quality issues are not systematically related to income, supporting its use as a control variable.

```{r eda distribution of share renewable}

## 3. Plot 1 — Verteilung share_renewable -------------------------------------
p1 <- muni_analysis %>%
  ggplot(aes(share_renewable)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", colour = "white") +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Distribution of renewable heating share",
    x     = "Share of buildings with renewable heating",
    y     = "Number of municipalities"
  ) +
  theme_minimal()

p1

```

### Distribution of share_renewable (Histogram)

The distribution shows a clear peak at 15–30%, with the highest concentration between 10% and 35%. Only a few municipalities exceed 50%, and the tail beyond 75% likely represents small alpine municipalities with a high share of wood heating. The histogram supports using the untransformed variable as the outcome, given that the skewness is moderate. A beta regression or an arcsine transformation would be indicated if there were a stronger concentration at the boundaries (near 0 or 1), which is not the case here.

```{r eda plot distribution of median income}

## 4. Plot 2 — Verteilung median_income ----------------------------------------
p2 <- muni_analysis %>%
  ggplot(aes(median_income)) +
  geom_histogram(binwidth = 5000, fill = "steelblue", colour = "white") +
  scale_x_continuous(labels = scales::comma_format(suffix = " CHF")) +
  labs(
    title = "Distribution of median taxable income",
    x     = "Median taxable income (CHF)",
    y     = "Number of municipalities"
  ) +
  theme_minimal()

p2

```

# Distribution of median_income (Histogram)

The income distribution is unimodal and approximately normally distributed in its core range (40,000–80,000 CHF), with a right-hand tail caused by low-tax municipalities. The sharp drop-off below 30,000 CHF reflects the Swiss price level; outliers below this threshold require methodological verification. For the regression, a robustness check using log-transformed income is recommended: the relationship between purchasing power and adoption behavior is plausibly non-linear (each additional franc has a stronger impact at lower income levels than at higher ones).

```{r eda scatterplot median income vs share renewable}

## 5. Plot 3 — Scatterplot median_income vs. share_renewable -------------------
typology_labels <- c("1" = "Urban", "2" = "Periurban", "3" = "Rural")

p3 <- muni_analysis %>%
  mutate(typology_3 = factor(typology_3, labels = typology_labels)) %>%
  ggplot(aes(median_income, share_renewable, colour = typology_3)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  scale_x_continuous(labels = scales::comma_format(suffix = " CHF")) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_colour_manual(values = c("steelblue", "darkorange", "forestgreen")) +
  labs(
    title   = "Median income vs. renewable heating share",
    x       = "Median taxable income (CHF)",
    y       = "Share of buildings with renewable heating",
    colour  = "Municipality type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p3

```

### Scatterplot: Income vs. Renewable Share by Municipality Type

The scatterplot yields the central finding of the EDA. Three patterns stand out:

    Direction: The relationship is negative across all three municipality types—as income increases, the share of renewable heating decreases. The OLS regression lines consistently display a negative slope, which is steepest for rural municipalities (green).

    Level Differences: At any given income level, rural municipalities (green) have a significantly higher renewable share than peri-urban (orange) and urban municipalities (blue). At around 55,000–65,000 CHF, rural municipalities sit approximately 15–25 percentage points above the urban level. This reflects a structural composition difference: wood heating and heat pumps in detached single-family homes dominate the rural building stock.

    Dispersion: The very wide dispersion—particularly in rural areas—shows that income alone explains only a fraction of the variance. This is consistent with r = –0.36. For the regression, this implies that further explanatory power must be derived from control variables and canton fixed effects.

Interpretation of the negative direction: The surprising negative sign is highly likely a composition effect: high-income municipalities are disproportionately urban (higher log_pop, more gas/district heating), while lower-income municipalities are disproportionately rural (more wood, heat pumps in single-family homes). Whether income exerts a true causal effect on heating adoption—independent of urbanity and building structure—will be revealed by the regression with controls.

```{r eda boxplot share renewable by typology}

## 6. Plot 4 — Boxplot share_renewable nach typology_3 ------------------------
p4 <- muni_analysis %>%
  mutate(typology_3 = factor(typology_3, labels = typology_labels)) %>%
  ggplot(aes(typology_3, share_renewable, fill = typology_3)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("steelblue", "darkorange", "forestgreen")) +
  labs(
    title = "Renewable heating share by municipality type",
    x     = NULL,
    y     = "Share of buildings with renewable heating"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p4

```

# Boxplot: Renewables by Municipality Type

The boxplot confirms the typological gradient: rural municipalities (median ~33%) > peri-urban (median ~23%) > urban (median ~15%). The IQR for rural municipalities (approx. 25–47%) is considerably wider than that of urban ones (~10–25%), indicating greater heterogeneity in rural areas—likely due to cantonal differences in subsidy regimes and local fuel availability (wood vs. heat pump potential).

Upper outliers are present across all typologies; for rural municipalities, they extend beyond 90%—representing individual alpine communities with historically established shares of wood heating. Urban outliers above 50% are rare and warrant an individual municipality-level inspection (e.g., municipalities with local heat pump programs or erroneous GWR entries).

```{r eda chloropleth maps}

# Nutze most_recent = TRUE, damit R die aktuellsten verfügbaren Karten zieht
gemeinden_sf <- bfs_get_base_maps(geom = "polg", most_recent = TRUE) 
kantone_sf   <- bfs_get_base_maps(geom = "kant", most_recent = TRUE) 

# Falls das klappt, kannst du direkt mit dem Join weitermachen:
map_data <- gemeinden_sf %>%
  mutate(id = as.numeric(id)) %>% 
  left_join(muni_analysis, by = c("id" = "bfs_id"))

# Gemeinsames Theme für beide Karten
map_theme <- theme_void(base_size = 11) +
  theme(
    legend.position      = "right",
    legend.key.height    = unit(1.2, "cm"),
    legend.key.width     = unit(0.35, "cm"),
    plot.title           = element_text(size = 12, face = "bold", margin = margin(b = 6)),
    plot.margin          = margin(5, 5, 5, 5)
  )

# Karte 1 — Anteil erneuerbare Heizungen
p_map_renewable <- ggplot() +
  geom_sf(data = map_data,
          aes(fill = share_renewable),
          colour = NA) +
  geom_sf(data = kantone_sf,
          fill = NA, colour = "white", linewidth = 0.5) +
  scale_fill_viridis_c(
    option   = "mako",
    direction = -1,
    labels   = scales::percent_format(accuracy = 1),
    na.value = "grey88",
    name     = "Share\nrenewable"
  ) +
  labs(title = "Renewable heating share by municipality (2024)") +
  map_theme

# Karte 2 — Medianeinkommen
p_map_income <- ggplot() +
  geom_sf(data = map_data,
          aes(fill = median_income),
          colour = NA) +
  geom_sf(data = kantone_sf,
          fill = NA, colour = "white", linewidth = 0.5) +
  scale_fill_viridis_c(
    option   = "rocket",
    direction = -1,
    labels   = scales::comma_format(suffix = " k", scale = 1/1000),
    na.value = "grey88",
    name     = "Median\nincome\n(CHF)"
  ) +
  labs(title = "Median taxable income by municipality (ESTV 2022)") +
  map_theme

p_map_renewable
p_map_income

```


## Limitations

Several limitations must be acknowledged. First, the analysis operates at the municipal aggregate level, meaning observed associations between income and renewable heating adoption cannot be attributed to individual households. The relevant financial capacity for the heating investment decisiInon is that of the building owner, not the average municipal resident. Second, the GWR heating entries are of limited reliability: as shown in Figure 1, approximately 50% of building entries are flagged as low quality. Third, the analysis does not account for cantonal subsidy heterogeneity beyond the fixed effects included in the model.