---
title: "Income and Renewable Heating Adoption in Switzerland"
author: "Yannik Zimmermann & Birk Diener"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(readr)
library(readxl)
library(ggplot2)
```

## Motivation

Switzerland has committed to achieving net-zero greenhouse gas emissions by 2050, with the heating sector representing one of the largest remaining sources of fossil fuel consumption. Transitioning residential buildings away from oil and gas heating towards renewable alternatives such as heat pumps, solar thermal systems, and biomass is therefore a critical lever of the energy transition. However, adoption rates of renewable heating systems vary significantly across Swiss municipalities (bfs 2024). This project investigates a potential explanation for this heterogenous adoption pattern, namely whether a statistically measurable relationship exists between community-level income and the share of renewable heating systems installed.

## Project Aim

This project aims to quantify the relationship between median income at the municipality level and the share of buildings equipped with renewable heating systems across Swiss municipalities. The analysis will address two core research questions: 

- Is there a statistically significant relationship between municipal-level financial capacity, measured through taxable income, and the share of residential buildings equipped with renewable heating systems across Swiss municipalities?
- To what extent is this relationship robust to structural characteristics of the building stock, namely the share of single-family houses and the age of the building stock, and to cantonal differences in subsidy programmes?

The analysis draws on the Federal Building and Dwellings Register (GWR) and ESTV municipal tax data. Given known data quality limitations in the GWR heating system entries, a systematic data recency assessment is integrated into the analysis prior to modelling. The findings will be visualised through choropleth maps and scatter plots to make spatial and socioeconomic patterns legible for a broader audience.

## Data Sources

Two datasets are used for this analysis, both openly available and joinable at the Swiss municipality level (BFS-Gemeindecode). The first is the Federal Building and Dwellings register (GWR), which records the primary heating energy, building category, and year of construction for every residential building in Switzerland. Key variables used in this analysis are the heating system type (GWAERZH1), the energy source (GENH1), the year of construction (GBAUJ), the information source of the heating entry (GWAERSCEH1), the last update date of the heating entry (GWAERDATH1), and the municipality code (GGDENR).

While the general availability of data to answer the research question is good, there is a significant challenge regarding the reliability of the heating system entries. The last comprehensive national update of this information took place as part of the Volkszählung 2000. Since then, municipal building departments are responsible for updating entries, primarily through building permit applications. As a result, data quality varies across municipalities, and heating systems installed without a building permit — such as most external air-source heat pumps — may not be recorded. We investigate this aspect further in the data preparation chapter.  

The second dataset is the income and wealth statistics from the Federal Tax Administration (ESTV), which provides median taxable income per municipality at the level of natural persons. It is published annually with a lag of approximately two years.

## Approach

The analysis proceeds in four steps: First, GWR building data is aggregated to the municipality level to compute the share of residential buildings equipped with a renewable heating system as a fraction of all buildings with a recorded heating system entry. Renewable heating is defined as heat pump, solar thermal, or biomass combined with energy source. District heating is excluded from the primary definition as its adoption is structurally determined by local infrastructure rather than individual investment decisions, but is included in a sensitivity analysis. The share of pre-1980 buildings per municipality serves as the building age control variable. 

Second, the municipal renewable heating share is joined to the ESTV income data using the BFS municipality code. Third, regression models are estimated using a stepwise approach: a baseline model with median income as the sole predictor, followed by successive addition of building structure controls (share of single-family houses, share of pre-1980 buildings), sociodemographic controls (municipal type, log population size), and cantonal fixed effects to absorb unobserved variation in subsidy programmes and building regulations. The municipal share of low-quality heating entries is included as a control variable throughout all model specifications. The stability of the income coefficient across specifications is reported. 

Fourth, findings are visualized through choropleth maps showing the spatial distribution of renewable heating adoption and median income respectively.

## Data Preparation
Before proceeding to the analysis, the GWR building data is prepared and aggregated to the municipal level. The preparation process includes filtering for residential buildings, flagging renewable heating systems, and assessing the quality of the heating system entries.

Residential buildings are identified using two GWR variables. First, GKAT (building category) is restricted to codes 1020 (exclusively residential) and 1030 (residential with secondary use), excluding buildings without mostly residential use. Second, GSTAT (building status) is restricted to code 1004, retaining only buildings recorded as currently existing and excluding buildings under construction, demolished, or otherwise inactive. Together, these two filters define the analytical population of existing residential buildings on which all subsequent measures are computed.

Heating systems are categorised based on the energy source variable (GENH1) across ten mutually exclusive groups: no heating system (7500), air-source heat pump (7501), ground- and water-source heat pump (7510–7513), gas (7520), oil (7530), wood and biomass (7540–7543), waste heat (7550), electricity (7560), solar thermal (7570), district heating (7580–7582), and unspecified or other (7598–7599). A building is flagged as renewable if its primary energy source is a heat pump, solar thermal, wood or biomass, waste heat, or recorded as having no heating system. The latter category is included on the grounds that buildings without an active heating system generate no fossil fuel consumption from space heating, making their exclusion from the renewable count methodologically conservative. District heating is retained as a separate flag for sensitivity analysis, as its adoption depends on local infrastructure availability rather than individual investment decisions.

```{r load gwr data}

## 1. Read the GWR building data --------------------------------------------
# The file is tab-separated even though it ends in .csv. We only read the
# columns we actually use (the full file has ~50 columns and ~3.3 M rows).

buildings <- read_tsv(
  here("gwr", "gebaeude_batiment_edificio.csv"),
  col_select = c(GGDENR, GGDENAME, GDEKT,
                 GKAT, GKLAS, GSTAT, GBAUJ,
                 GWAERZH1, GENH1, GWAERSCEH1, GWAERDATH1),
  na = c("", "NA"),
  show_col_types = FALSE
)

## 2. Keep only existing residential buildings -----------------------------
# GKAT: 1020 = exclusively residential, 1030 = residential with side use
# GSTAT 1004 = the building actually exists today.

residential <- buildings %>%
  filter(GKAT %in% c(1020, 1030), GSTAT == 1004)


## 3. Flag each building --------------------------------------------------
# Renewable heating per the project brief:

residential <- residential %>%
  mutate(
    # Kein Heizsystem — 7500
    is_no_heatsystem = GENH1 == 7500,
    # Wärmepumpe Luft — 7501
    is_hp_air        = GENH1 == 7501,
    # Wärmepumpe Erde/Wasser — 751x
    is_hp_ground     = GENH1 %in% c(7510, 7511, 7512, 7513),
    # Fossil — 752x, 753x
    is_gas           = GENH1 == 7520,
    is_oil           = GENH1 == 7530,
    # Holz — 754x
    is_wood          = GENH1 %in% c(7540, 7541, 7542, 7543),
    # Abwärme — 755x
    is_waste_heat    = GENH1 == 7550,
    # Elektrizität — 756x
    is_electric      = GENH1 == 7560,
    # Solar thermisch — 757x
    is_solar         = GENH1 == 7570,
    # Fernwärme — 758x (Sensitivity Analysis)
    is_district      = GENH1 %in% c(7580, 7581, 7582),
    # Unbestimmt/Andere — 759x
    is_other         = GENH1 %in% c(7598, 7599),
    
    # Aggregierte Flags
    is_heatpump  = is_hp_air | is_hp_ground,
    is_renewable = is_heatpump | is_wood | is_solar | 
                   is_waste_heat | is_no_heatsystem
  )

## 5. Aggregate to the municipality ---------------------------------------
# Renewable share is defined over buildings with a heating entry, so the
# denominator is sum(!is.na(GWAERZH1)), not n().

muni <- residential %>%
  group_by(bfs_id = GGDENR, municipality = GGDENAME, canton = GDEKT) %>%
  summarise(
    n_buildings     = n(),
    n_with_heating  = sum(!is.na(GWAERZH1)),
    renewable_share = sum(is_renewable)   / sum(!is.na(GWAERZH1)),
    .groups = "drop"
  )
```


Unfortunately, the heating system entries in the GWR are of limited reliability. The last comprehensive national update of this information took place as part of the Volkszählung 2000, and since then, municipal building departments are responsible for updating entries, primarily through building permit applications. As a result, data quality varies across municipalities, and heating systems installed without a building permit — such as most external air-source heat pumps — may not be recorded.
This is visible in Figure 1, which plots the most recent update year of the heating entry per building: approximately 600,000 residential buildings have not had their heating entry updated since 2000, and a further 1.08 million buildings have no update date recorded at all. Together these two groups represent approximately 50% of the 3.3 million buildings in the GWR.

```{r gwr data quality}

## 4. Quick data-quality check --------------------------------------------
# When was the heating entry last updated? The brief notes this is a problem,
# many entries still go back to the 2000 census.

residential %>%
  mutate(year = as.integer(substr(GWAERDATH1, 1, 4))) %>%
  filter(!is.na(year), year >= 1990) %>%
  ggplot(aes(year)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Latest update of the GWR heating entry",
       x = NULL, y = "Number of buildings") +
  theme_minimal()

```

Figure 1: Distribution of GWR heating entry update years for residential buildings in Switzerland

*Note: The figure shows the number of residential buildings by the year of their most recent heating system entry update (GWAERDATH1). Buildings with no recorded update date (n = 1,083,361) are excluded from the plot. The spike in 2001 reflects entries sourced from the Volkszählung 2000 (GWAERSCEH1 = 860), which represent the last comprehensive national update of heating system data. The subsequent increase from 2019 onwards reflects growing administrative activity following the 2017 revision of the GWR ordinance, which made heating data recording mandatory for new building permits.*

To account for this, each building is assigned a binary low-quality flag if its heating entry either has no recorded update date or is explicitly sourced from the Volkszählung 2000 (GWAERSCEH1 = 860). Aggregated to the municipal level, this produces a share of low-quality heating entries per municipality, which is used as a control variable in the regression models. The median municipality has 58% of its heating entries flagged as low quality, with an interquartile range of 48.5% to 69.3%, indicating that the data quality challenge is structural rather than isolated. Results should therefore be interpreted as conservative estimates: low-quality entries are likely concentrated in older building stock that is disproportionately fossil-heated, meaning the measured renewable heating shares likely underestimate true adoption rates, and any detected income effect is likely attenuated by this noise rather than inflated.


```{r load estv data}

estv_file <- here("estv", "statistik-dbst-np-gden-2022-normalfall.xlsx")

read_total <- function(sheet, value_name) {
  raw <- read_excel(
    estv_file, sheet = sheet, skip = 2,
    col_types = c("text", "text", "numeric", "text", rep("numeric", 11)),
    na = c("", "NA", "-")                    # <- '-' ergänzt
  )
  names(raw)[1:4]       <- c("canton_id", "canton_name", "bfs_id", "muni_name")
  names(raw)[ncol(raw)] <- "total"
  raw %>%
    filter(!is.na(bfs_id)) %>%
    select(bfs_id, !!value_name := total)
}

counts <- read_total("511", "n_taxpayers")
sums   <- read_total("512", "sum_taxable")

income <- counts %>%
  left_join(sums, by = "bfs_id") %>%
  mutate(mean_taxable = sum_taxable / n_taxpayers)

```

To prepare the data for the regression analysis, the building-level flags are extended with a low-quality marker, a single-family-house flag, and a pre-1980 construction flag. The municipal aggregation is then re-computed to add three structural control variables (`low_quality_share`, `sfh_share`, `pre1980_share`) and an alternative renewable share that includes district heating buildings in the numerator (`renewable_share_dh`) for the sensitivity analysis. The aggregated municipal table is joined with the ESTV income data on the BFS municipality code. Log transformations are applied to the mean taxable income (`log_income`) and to the building count (`log_pop`, used as a proxy for municipality size since no separate population variable is available from the GWR). Municipalities are further classified into three size groups (`rural`, `semi-urban`, `urban`) based on tertiles of `n_buildings`. To improve readability of all tables and plots, canton numeric codes are mapped to their standard two-letter abbreviations.

```{r prepare analysis data, warning=FALSE, message=FALSE}

## 1. Additional building-level flags --------------------------------------
# Low quality: heating entry is from the 2000 census (GWAERSCEH1 == 860) or
# has no recorded update date at all (is.na(GWAERDATH1)).
# SFH: GKLAS == 1110 (single-family house).
# Pre-1980: construction year before 1980. NAs in GBAUJ count as FALSE.

residential <- residential %>%
  mutate(
    is_lowqual = GWAERSCEH1 %in% 860 | is.na(GWAERDATH1),
    is_sfh     = !is.na(GKLAS) & GKLAS == 1110,
    is_pre1980 = !is.na(GBAUJ) & GBAUJ < 1980
  )

## 2. Re-aggregate to the municipality with all controls ------------------

muni <- residential %>%
  group_by(bfs_id = GGDENR, municipality = GGDENAME, canton_id = GDEKT) %>%
  summarise(
    n_buildings        = n(),
    n_with_heating     = sum(!is.na(GWAERZH1)),
    renewable_share    = sum(is_renewable)                 / sum(!is.na(GWAERZH1)),
    renewable_share_dh = sum(is_renewable | is_district)   / sum(!is.na(GWAERZH1)),
    low_quality_share  = sum(is_lowqual)                   / sum(!is.na(GWAERZH1)),
    sfh_share          = sum(is_sfh)                       / n(),
    pre1980_share      = sum(is_pre1980)                   / n(),
    .groups = "drop"
  )
# `muni` now has one row per municipality, with the renewable share, the
# alternative renewable share including district heating, and the three
# control variables for the regression.

## 3. Rename canton column ------------------------------------------------
# GDEKT already contains two-letter canton abbreviations (ZH, BE, etc.)

muni <- muni %>%
  rename(canton = canton_id)

## 4. Join with ESTV income, add log / categorical variables --------------

df <- muni %>%
  left_join(income %>% select(bfs_id, mean_taxable), by = "bfs_id") %>%
  mutate(
    log_income = log(mean_taxable),
    log_pop    = log(n_buildings),
    muni_type  = factor(
      ntile(n_buildings, 3),
      levels = 1:3,
      labels = c("rural", "semi-urban", "urban")
    )
  )
# `df` is the analysis frame: one row per municipality with renewable_share
# (and renewable_share_dh), the income predictor (log_income), and all controls.

## 5. Filter very small municipalities and missing predictors -------------
# Municipalities with fewer than 20 buildings with a heating entry produce
# unstable shares and are dropped, as are rows missing income or share.

n_before <- nrow(df)
df <- df %>%
  filter(n_with_heating >= 20,
         !is.na(log_income),
         !is.na(renewable_share)) %>%
  mutate(
    # Recreate muni_type based on post-filtered data to ensure all levels have observations
    muni_type = factor(
      ntile(n_buildings, 3),
      levels = 1:3,
      labels = c("rural", "semi-urban", "urban")
    ),
    canton = factor(canton)
  )
n_after   <- nrow(df)
n_dropped <- n_before - n_after

```

After filtering, `r n_after` municipalities are retained out of `r n_before` (`r n_dropped` dropped, predominantly small alpine or commuter municipalities with fewer than 20 buildings recorded with a heating entry).

## Exploratory Visualisations

Before estimating any regression model, the renewable heating share is inspected visually. Three plots characterise the data at the municipality level (distribution, relationship to income, relationship to data quality), and two choropleth maps show the spatial pattern of renewable adoption and mean taxable income across Switzerland.

```{r fig-2, fig.cap="Figure 2: Distribution of the renewable heating share across Swiss municipalities. The unit is the fraction of residential buildings with a recorded heating entry. The dashed red line marks the national mean across municipalities."}

p2 <- ggplot(df, aes(x = renewable_share)) +
  geom_histogram(bins = 40, fill = "steelblue", colour = "white") +
  geom_vline(aes(xintercept = mean(renewable_share, na.rm = TRUE)),
             linetype = "dashed", colour = "firebrick") +
  labs(x = "Renewable heating share",
       y = "Number of municipalities") +
  theme_minimal()
p2

```

```{r fig-3, fig.cap="Figure 3: Mean taxable income (log) vs. renewable heating share. Each dot is one municipality, coloured by canton. The black line is an OLS fit across all municipalities."}

p3 <- ggplot(df, aes(x = log_income, y = renewable_share)) +
  geom_point(aes(colour = canton), alpha = 0.6, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(x = "Log mean taxable income (CHF per taxpayer)",
       y = "Renewable heating share",
       colour = "Canton") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2, override.aes = list(size = 2.5)))
p3

```

```{r fig-4, fig.cap="Figure 4: Share of low-quality heating entries vs. renewable heating share. The negative association illustrates the data-quality confound: municipalities with more outdated GWR entries appear to have lower renewable adoption."}

p4 <- ggplot(df, aes(x = low_quality_share, y = renewable_share)) +
  geom_point(aes(colour = canton), alpha = 0.6, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(x = "Share of low-quality heating entries",
       y = "Renewable heating share",
       colour = "Canton") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 2, override.aes = list(size = 2.5)))
p4

```

```{r map-data, warning=FALSE, message=FALSE}

library(sf)

# Swiss municipality polygons. The cached file is produced from the swisstopo
# swissBOUNDARIES3D dataset and stored under data-derived/municipalities.gpkg.
# If the shapefile's BFS code variable is named differently, adjust the join
# key (currently expected: `bfs_id`).

muni_polygons <- st_read(
  here("data-derived", "municipalities.gpkg"),
  quiet = TRUE
) %>%
  st_zm()

map_data <- muni_polygons %>%
  left_join(df, by = "bfs_id")
# `map_data` is the sf object containing geometry plus all df columns.

```

```{r fig-5, fig.cap="Figure 5: Choropleth map of the municipal renewable heating share. Municipalities dropped from the analysis appear grey."}

p5 <- ggplot(map_data) +
  geom_sf(aes(fill = renewable_share), colour = NA) +
  scale_fill_viridis_c(option = "plasma",
                       labels = scales::label_percent(),
                       na.value = "grey90") +
  coord_sf() +
  labs(fill = "Renewable share") +
  theme_void()
p5

```

```{r fig-6, fig.cap="Figure 6: Choropleth map of the municipal mean taxable income (log scale). Higher values indicate higher mean taxable income per taxpayer."}

p6 <- ggplot(map_data) +
  geom_sf(aes(fill = log_income), colour = NA) +
  scale_fill_viridis_c(option = "plasma",
                       na.value = "grey90") +
  coord_sf() +
  labs(fill = "Log mean taxable income") +
  theme_void()
p6

```

## Analysis

The analysis follows a deliberate funnel: first describe the variables on their own, then estimate the main relationship in a series of nested regressions that progressively rule out alternative explanations, then stress-test the result by changing the definition of the dependent variable, and finally check that the OLS assumptions hold. The motivation is that a naive correlation between municipal income and renewable adoption could reflect almost anything that comes bundled with income (newer housing, more owner-occupiers, urban location, generous cantonal subsidies), so each modelling step is designed to isolate the part of the relationship that survives the addition of a specific group of controls.

### Descriptive Statistics

```{r descriptive-stats, warning=FALSE, message=FALSE}

library(tidyr)

desc_stats <- df %>%
  select(renewable_share, log_income, low_quality_share, sfh_share, pre1980_share) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(
    mean   = mean(value,   na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd     = sd(value,     na.rm = TRUE),
    min    = min(value,    na.rm = TRUE),
    max    = max(value,    na.rm = TRUE),
    .groups = "drop"
  )
# `desc_stats` is a tidy summary table of central tendency and spread.

knitr::kable(desc_stats, digits = 2,
             caption = "Table 1: Summary statistics for the key analysis variables across Swiss municipalities.")

```

Before fitting any model, the analysis variables are summarised on their own as a sanity check: do any of them have a degenerate distribution, an unexpected magnitude, or a spread so narrow that no regression could pick up a signal? The mean renewable heating share across municipalities is approximately `r round(mean(df$renewable_share, na.rm = TRUE)*100, 1)`%, with a wide spread (sd = `r round(sd(df$renewable_share, na.rm = TRUE), 2)`), confirming the substantial heterogeneity across municipalities that motivates the project in the first place. The log mean taxable income has a median of about `r round(median(df$log_income, na.rm = TRUE), 2)` (roughly CHF `r format(round(exp(median(df$log_income, na.rm = TRUE))), big.mark = "'")` per taxpayer), with a comparatively narrow distribution: most municipalities sit close to the centre and only a thin tail of high-income outliers stretches the upper end. Neither variable is degenerate, so both can carry information for the regression that follows.

### Stepwise Regression

The relationship between municipal income and renewable heating share is estimated using ordinary least squares (OLS) in four nested specifications. The reason for fitting more than one model is that a single regression of `renewable_share` on `log_income` could conflate income with everything that comes bundled with it: newer building stock, a higher share of owner-occupied single-family houses, more urban location, more generous cantonal subsidy regimes. Each successive model therefore adds one group of controls and asks: is the income effect really about *this*?

Model 1 is the baseline: only income and the `low_quality_share` data-quality control. Model 2 adds the building stock controls (share of single-family houses, share of pre-1980 buildings); if the income coefficient shrinks here, part of the original effect was really "richer municipalities have newer or more owner-friendly housing". Model 3 adds municipal size and the rural / semi-urban / urban classification, asking whether the effect is really "richer municipalities are urban". Model 4 adds canton as a factor, which restricts comparisons to municipalities *within the same canton* and so absorbs cantonal subsidy regimes, building regulations, climate, and anything else that is constant within a canton. The `low_quality_share` control is carried through all four models to partly absorb the GWR data quality confound discussed earlier. If the income coefficient survives each step, the association is robust to that group of confounders; if it collapses at a particular step, that step has identified the real driver.

```{r stepwise-regression, warning=FALSE, message=FALSE}

library(broom)

model1 <- lm(renewable_share ~ log_income + low_quality_share,
             data = df)

model2 <- lm(renewable_share ~ log_income + low_quality_share +
               sfh_share + pre1980_share,
             data = df)

model3 <- lm(renewable_share ~ log_income + low_quality_share +
               sfh_share + pre1980_share + muni_type + log_pop,
             data = df)

model4 <- lm(renewable_share ~ log_income + low_quality_share +
               sfh_share + pre1980_share + muni_type + log_pop + canton,
             data = df)

# Helper: format a p-value with conventional significance stars
# (*** < 0.001, ** < 0.01, * < 0.05, . < 0.1) and at least 3 decimal places.
fmt_p <- function(p) {
  stars <- ifelse(p < 0.001, "***",
           ifelse(p < 0.01,  "**",
           ifelse(p < 0.05,  "*",
           ifelse(p < 0.1,   ".", ""))))
  txt <- ifelse(p < 0.001, "<0.001", formatC(p, digits = 3, format = "f"))
  trimws(paste(txt, stars))
}

# Helper: pull the income coefficient + model R^2 out of one model.
extract_income <- function(model, name) {
  tidy(model) %>%
    filter(term == "log_income") %>%
    transmute(
      Model       = name,
      `Beta`      = estimate,
      SE          = std.error,
      `p-value`   = fmt_p(p.value),
      `R-squared` = glance(model)$r.squared
    )
}

income_table <- bind_rows(
  extract_income(model1, "1: Baseline"),
  extract_income(model2, "2: + Building structure"),
  extract_income(model3, "3: + Sociodemographic"),
  extract_income(model4, "4: + Cantonal FE")
)
# `income_table` summarises the income coefficient across the four model specs.

knitr::kable(income_table, digits = 4,
             caption = "Table 2: The income coefficient (Beta on log_income) across the four model specifications, with standard errors, p-values, and overall model R-squared. Significance: *** p<0.001, ** p<0.01, * p<0.05, . p<0.1.")

```

To read the coefficient: a 1-unit change in `log_income` corresponds to roughly a 2.7-fold change in real income (since exp(1) is approximately 2.72), or, framed in everyday terms, a 1% change in income is associated with a change in the predicted renewable share of about one one-hundredth of the reported coefficient. A coefficient of 0.05 therefore implies that a 2.7-fold income increase is associated with a 5-percentage-point higher renewable heating share, holding the controls constant.

Reading the table down the column, the income coefficient keeps the same sign and statistical significance across all four specifications. It attenuates from Model 1 to Model 4, which is exactly what one would expect: each set of added controls (building stock, municipal type, cantonal fixed effects) is itself correlated with income and so absorbs part of the raw gradient. The substantively important observation is that the coefficient does not collapse to zero, and does not lose significance once cantonal fixed effects are added. This means the income-renewable association is not a pure artefact of richer cantons happening to subsidise more: some portion of the gradient is genuinely *within-canton*, that is, visible when comparing a richer municipality to a poorer one in the same cantonal subsidy regime.

### Sensitivity Analysis: Including District Heating

Excluding district heating from the renewable definition is a judgement call. The reasoning was that district-heating availability is determined by local infrastructure rather than by individual investment decisions, so including it would risk mistaking infrastructure for wealth. But the call is debatable, so the same Model 4 is re-estimated with `renewable_share_dh` (renewable share including district heating) as the dependent variable. If the income coefficient stays roughly the same, the substantive conclusion is robust to the definition; if it changes materially, the definitional choice is doing more work than the data.

```{r sensitivity, warning=FALSE, message=FALSE}

model_sens <- lm(renewable_share_dh ~ log_income + low_quality_share +
                   sfh_share + pre1980_share + muni_type + log_pop + canton,
                 data = df)

sens_row <- tidy(model_sens) %>%
  filter(term == "log_income") %>%
  transmute(
    Model       = "Model 4 + district heating",
    `Beta`      = estimate,
    SE          = std.error,
    `p-value`   = fmt_p(p.value),
    `R-squared` = glance(model_sens)$r.squared
  )

knitr::kable(sens_row, digits = 4,
             caption = "Table 3: Sensitivity analysis. Model 4 re-estimated with district heating included in the renewable share.")

```

Including district heating in the numerator changes the income coefficient only marginally relative to the primary Model 4 estimate. This indicates that the income-renewable relationship documented above is not driven by district-heating municipalities, which are typically larger and more urban, but reflects the adoption of individual renewable systems (heat pumps, wood, solar thermal) across the full sample.

### Model Diagnostics

OLS assumes that the residuals have mean zero, constant variance, and are approximately normally distributed. Violations don't automatically invalidate the point estimates but they distort the standard errors and therefore the p-values, which is what matters for the significance claims made above. Three quick diagnostics are run on Model 4 to flag any obvious violations. The residuals-vs-fitted plot tests the first two assumptions: a flat cloud around zero with no funnel and no curvature is what we want. The Q-Q plot tests normality: the dots should hug the reference line. The count of studentised residuals beyond an absolute value of three is a sniff test for whether a small number of unusual municipalities might be dragging the fit.

```{r diagnostics, warning=FALSE, message=FALSE, fig.cap="Figure 7: Model 4 diagnostic plots. Left: residuals vs. fitted values. Right: normal Q-Q plot of residuals."}

diag_df <- tibble(
  fitted = fitted(model4),
  resid  = resid(model4),
  rstud  = rstudent(model4)
)

p_resid <- ggplot(diag_df, aes(fitted, resid)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "firebrick") +
  geom_smooth(se = FALSE, colour = "steelblue") +
  labs(title = "Residuals vs fitted (Model 4)",
       x = "Fitted values", y = "Residuals") +
  theme_minimal()

p_qq <- ggplot(diag_df, aes(sample = resid)) +
  stat_qq(alpha = 0.4) +
  stat_qq_line(colour = "firebrick") +
  labs(title = "Q-Q plot of residuals (Model 4)",
       x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal()

p_resid
p_qq

n_outliers <- sum(abs(diag_df$rstud) > 3, na.rm = TRUE)

```

A total of `r n_outliers` municipalities have an absolute studentised residual greater than 3. The residuals vs. fitted plot shows no severe systematic pattern around zero, and the Q-Q plot indicates approximate normality of residuals, with mild deviation in the tails that is consistent with the heterogeneity expected at this aggregation level.

## Limitations

Several limitations must be acknowledged. First, the analysis operates at the municipal aggregate level, meaning observed associations between income and renewable heating adoption cannot be attributed to individual households. The relevant financial capacity for the heating investment decisiInon is that of the building owner, not the average municipal resident. Second, the GWR heating entries are of limited reliability: as shown in Figure 1, approximately 50% of building entries are flagged as low quality. Third, the analysis does not account for cantonal subsidy heterogeneity beyond the fixed effects included in the model.

## Conclusion

This project investigated whether municipal-level taxable income predicts the share of buildings equipped with renewable heating systems across Swiss municipalities, using the Federal Building and Dwellings Register (GWR) and ESTV income data. After joining both sources at the BFS municipality level and adjusting for the data quality of the GWR heating entries, building stock structure, municipal size, and cantonal fixed effects, a statistically significant positive association between municipal income and the renewable heating share is observed. In the most complete specification (Model 4), the estimated coefficient on `log_income` is `r round(coef(model4)["log_income"], 3)`, meaning that municipalities with higher mean taxable income tend to also have higher renewable heating adoption, holding the controls constant.

This association is plausibly multi-causal rather than purely financial. Higher-income municipalities tend to have a larger share of owner-occupied housing, with owners who have both the capital to invest in heat pumps or solar thermal systems and the incentive to do so. They also tend to have somewhat newer building stock, which is more compatible with heat-pump installation, and may exhibit stronger demand for green upgrades for reputational or quality-of-life reasons. The observational design of this study does not, however, allow these mechanisms to be disentangled, and the association should not be interpreted as a causal income effect.

A second important caveat concerns data quality. The `low_quality_share` control does not fully eliminate attenuation bias from stale or missing GWR entries. Since renewable systems (in particular external air-source heat pumps) are systematically underrepresented in older entries, the true income effect on actual renewable adoption is likely stronger than the estimate reported here.

From a policy perspective, the finding suggests that means-tested subsidies could play a useful role in offsetting the income gradient if the goal is to make renewable heating uptake more uniform across municipalities. Future work should ideally combine individual building-level panel data with cantonal subsidy amounts as a continuous covariate, and apply spatial autoregressive models to account for the spatial clustering of adoption visible in the choropleth maps. A targeted re-survey of low-quality GWR entries in a representative sample of municipalities would also substantially improve the credibility of any quantitative claim about the income gradient.

<!-- BONUS CHAPTER: optional -->

## Additional Chapter: Cantonal Deep-Dive (Optional)

*This chapter is optional and provides an exploratory deep-dive into three cantons known to maintain high GWR data quality: Luzern (LU), Basel-Landschaft (BL), and Basel-Stadt (BS). The motivation is that the income-renewable relationship may be estimated more cleanly in these cantons, where heating entries are more reliable.*

### A1: Data Quality Profile of LU, BL, BS vs. Switzerland

```{r bonus-a1, warning=FALSE, message=FALSE, fig.cap="Figure 8: Distribution of the share of low-quality heating entries by group. Lower values indicate better data quality. LU, BL, and BS are compared to the rest of Switzerland."}

focus_cantons <- c("LU", "BL", "BS")

df_box <- df %>%
  mutate(group = case_when(
    canton %in% focus_cantons ~ canton,
    TRUE                      ~ "Rest of CH"
  )) %>%
  mutate(group = factor(group, levels = c("Rest of CH", "LU", "BL", "BS")))
# `df_box` adds a grouping variable for the boxplot.

p_bonus1 <- ggplot(df_box, aes(x = group, y = low_quality_share, fill = group)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.4) +
  scale_fill_viridis_d() +
  labs(x = NULL, y = "Share of low-quality heating entries") +
  theme_minimal() +
  theme(legend.position = "none")
p_bonus1

```

The boxplot allows a visual comparison of GWR data quality across the four groups. The three focus cantons should display systematically lower medians and tighter interquartile ranges than the rest of Switzerland if the premise of cleaner cantonal record-keeping holds.

### A2: Income-Renewable Relationship in Focus Cantons

```{r bonus-a2, warning=FALSE, message=FALSE, fig.cap="Figure 9: Renewable heating share vs. log mean taxable income, faceted by focus canton (LU, BL, BS). Each dot is one municipality; the red line is an OLS fit within canton."}

df_focus <- df %>% filter(canton %in% focus_cantons)
# `df_focus` is the LU/BL/BS subset for the focused regression.

model_bonus <- lm(renewable_share ~ log_income + low_quality_share +
                    sfh_share + pre1980_share,
                  data = df_focus)

focus_row <- tidy(model_bonus) %>%
  filter(term == "log_income") %>%
  transmute(
    Model       = "Focus cantons only (LU/BL/BS, Model 2)",
    `Beta`      = estimate,
    SE          = std.error,
    `p-value`   = fmt_p(p.value),
    `R-squared` = glance(model_bonus)$r.squared
  )

knitr::kable(
  bind_rows(extract_income(model2, "National (Model 2)"), focus_row),
  digits = 4,
  caption = "Table 4: Income coefficient in the national Model 2 vs. the focus-canton subset."
)

p_bonus2 <- ggplot(df_focus, aes(x = log_income, y = renewable_share)) +
  geom_point(alpha = 0.6, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "firebrick") +
  facet_wrap(~ canton) +
  labs(x = "Log mean taxable income (CHF per taxpayer)",
       y = "Renewable heating share") +
  theme_minimal()
p_bonus2

```

The focus-canton regression estimates the relationship in a subsample with structurally better GWR data quality. If the national estimate is attenuated by data noise, the focus-canton coefficient should be larger in absolute terms. The faceted scatter plot shows the same relationship visually within each of the three cantons.

### A3: Analogue Canton Identification

```{r bonus-a3, warning=FALSE, message=FALSE}

# Canton-level medians on observable structural dimensions.
canton_profile <- df %>%
  group_by(canton) %>%
  summarise(
    sfh_med        = median(sfh_share,        na.rm = TRUE),
    pre1980_med    = median(pre1980_share,    na.rm = TRUE),
    log_income_med = median(log_income,       na.rm = TRUE),
    lowqual_med    = median(low_quality_share, na.rm = TRUE),
    .groups = "drop"
  )

# Standardise each column (z-scores) so dimensions are comparable.
profile_mat <- as.matrix(canton_profile[, -1])
profile_mat <- scale(profile_mat)
rownames(profile_mat) <- canton_profile$canton

dist_mat <- as.matrix(dist(profile_mat, method = "euclidean"))
# `dist_mat` is a square matrix of pairwise z-score distances between cantons.

# For each focus canton, list the top 5 nearest neighbours (excluding itself
# and the other two focus cantons).
nearest_table <- bind_rows(lapply(focus_cantons, function(fc) {
  dists <- dist_mat[fc, ]
  others <- dists[!names(dists) %in% c(fc)]
  top5   <- sort(others)[1:5]
  tibble(
    `Focus canton`       = fc,
    `Nearest neighbour`  = names(top5),
    `Distance (z-score)` = as.numeric(top5)
  )
}))

knitr::kable(nearest_table, digits = 3,
             caption = "Table 5: Top 5 structurally nearest cantons to each focus canton, by Euclidean distance in z-score space.")

# Closest one analogue per focus canton (excluding the other focus cantons),
# pooled into a unique set for the joint regression.
analogues <- unique(unlist(lapply(focus_cantons, function(fc) {
  d <- dist_mat[fc, ]
  d <- d[!names(d) %in% focus_cantons]
  names(sort(d))[1]
})))

df_focus_plus <- df %>% filter(canton %in% c(focus_cantons, analogues))

model_bonus_plus <- lm(renewable_share ~ log_income + low_quality_share +
                         sfh_share + pre1980_share,
                       data = df_focus_plus)

comparison <- bind_rows(
  extract_income(model2, "National (Model 2)"),
  tidy(model_bonus) %>%
    filter(term == "log_income") %>%
    transmute(Model       = "Focus cantons only",
              `Beta`      = estimate,
              SE          = std.error,
              `p-value`   = fmt_p(p.value),
              `R-squared` = glance(model_bonus)$r.squared),
  tidy(model_bonus_plus) %>%
    filter(term == "log_income") %>%
    transmute(Model       = sprintf("Focus + analogues (%s)",
                                     paste(analogues, collapse = ", ")),
              `Beta`      = estimate,
              SE          = std.error,
              `p-value`   = fmt_p(p.value),
              `R-squared` = glance(model_bonus_plus)$r.squared)
)
# `comparison` lines up the income coefficient across three nested samples.

knitr::kable(comparison, digits = 4,
             caption = "Table 6: Income coefficient comparison across national, focus-canton-only, and focus-plus-analogue samples (all using Model 2 specification).")

```

If the focus-canton estimate is similar to the focus-plus-analogue estimate but larger than the national one, this supports the interpretation that data quality is the main driver of attenuation in the national sample. Conversely, if the focus-canton coefficient is closer to the national estimate once analogues are added, this would suggest that the focus cantons are not representatively cleaner but simply different in other observable ways. The interpretation should remain cautious: with only three focus cantons and one or two analogues each, the sample size is small and standard errors are correspondingly wide.